---
output:
  pdf_document: default
  output: rticles::ctex
  html_notebook: default
author: "Nkonzo Sithole"
date: '`r Sys.Date()`'
title: "Employee Attrition Prediction"
documentclass: ctexart
---
  
1. **Introduction**

Employees are the most important aset to organisations, hiring and retention of top talent is an extremely challenging task that requires capital, time and skills.

Employee attrition generally has negatively impact to many companies. Companies must have an HR strategy about hiring and retention, I have personally observer that many companies have internal surveys to check where they can improve to assist to prepare or avoid for such loss.
  
For example, studies found that staff churn is correlated with both demographic information as well as behavioral activities, satisfaction, etc. 

I will be looking to predictors that must be taken into consideration by companies. Machine learning models or techniques can give better prediction on employee attrition, as by nature they mathematically model the correlation between factors and attrition outcome and maximize 

  
In this study, (https://towardsdatascience.com/employee-retention-using-machine-learning-e7193e84bec4), they were looking at the cause of such leaving.
I will use the data from  https://www.kaggle.com/pavansubhasht/ibm-hr-analytics-attrition-dataset


**libraries**
```{r}

library("dplyr") 
library("ggplot2")
library("RColorBrewer")
library("plotrix")
library("forcats")
library("ggplot2")
library("caret")
library("corrplot")
library("corrgram")
library("gridExtra")
library("grid")

```
2. **Data exploration**

The experiments will be conducted on a data set of employees. The data set is publicly available and can be found at https://www.kaggle.com/pavansubhasht/ibm-hr-analytics-attrition-dataset.
```{r}
data <- read.csv("WA_Fn-UseC_-HR-Employee-Attrition.csv")
head(data)
```
## Summary
```{r}
summary(data)
```
## Fix Age column
```{r}
colnames(data)[1] <- "Age"
```
## Dataset is made up of the following rows and columns
```{r}
str(data)
cat("Data Set has ",dim(data)[1], " Rows and ", dim(data)[2], " Columns" )
```
## Checking missing and duplicate values
```{r}
sum(is.na(duplicated(data)))
```
3. **Data Visualization**

```{r}
data %>%
  group_by(Attrition) %>%
  tally() %>%
  ggplot(aes(x = Attrition, y = n,fill=Attrition)) +
  geom_bar(stat = "identity") +
  theme_minimal()+
  labs(x="Attrition", y="Count of Attriation")+
  ggtitle("Attrition")+
  geom_text(aes(label = n), vjust = -0.5, position = position_dodge(0.9))
```
b) Checking employees status(attrition) per job title
```{r}
ggplot(data, aes(JobRole, fill=Attrition)) +
  geom_bar(aes(y=(..count..)/sum(..count..)), position="dodge") +
  xlab("Job Role") +
  ylab("Percentage")
```

c) Income, jobRole, previous percentage salary hike and service years may affect decision for employees to leave. 

```{r}
ggplot(filter(data, (PercentSalaryHike >= 11) & (YearsAtCompany >= 2) & (YearsAtCompany <= 5) & (JobLevel < 3)),
       aes(x=factor(JobRole), y=MonthlyIncome, color=factor(Attrition))) +
  geom_boxplot() +
  xlab("Department") +
  ylab("Monthly income") +
  scale_fill_discrete(guide=guide_legend(title="Attrition")) +
  theme_bw() +
  theme(text=element_text(size=13), legend.position="top")
```

d) Employees grid graph in relation with Years of service, Growth, Manager, Income and salary increase.
```{r}
EmployeesYearOfService <- ggplot(data,aes(YearsAtCompany,fill = Attrition))+geom_bar()
EmployeesGrowth <- ggplot(data,aes(YearsSinceLastPromotion,fill = Attrition))+geom_bar()
EmployeesManager <- ggplot(data,aes(YearsWithCurrManager,fill = Attrition))+geom_bar()
EmployeeSalIncrease <- ggplot(data,aes(PercentSalaryHike,Attrition))+geom_point(size=4,alpha = 0.01)
EmployeesIncome <- ggplot(data,aes(MonthlyIncome,fill=Attrition))+geom_density()
gr <- grid.arrange(EmployeesYearOfService,EmployeesGrowth,EmployeesManager,EmployeeSalIncrease,EmployeesIncome,ncol=2,top = "Grid graphs")
```

```{r}
gr
```

## data correlation

## remove near zero variables

```{r}
near_Zero_variables <- names(data[, nearZeroVar(data)]) %>% print()
data <- data %>% select(-one_of(near_Zero_variables))
```
corrgram(data, lower.panel = panel.shade, upper.panel = panel.pie, text.panel = panel.txt, main = "Corrgram of all numeric  variables")

From this,  I will use algorithms like rf or XGBoost to build a model that can predict in fact which employees are most likely to leave in the future 

4. **Data Preparation and Partitioning** 

## convert certain integer variable to factor variable.

```{r}
factor_variables <- c("Education", "EnvironmentSatisfaction", "JobInvolvement", "JobLevel", "JobSatisfaction", "NumCompaniesWorked", "PerformanceRating", "RelationshipSatisfaction", "StockOptionLevel")
data[, factor_variables] <- lapply((data[, factor_variables]), as.factor)
data <- data %>% mutate_if(is.character, as.factor)
str(data)
```

Before modeling, first I use `set.seed(1)` and partition my data into train and test sets, which will be used to model and produce predictions. 
Then towards the end of this report, I will show the final model performance on the validation set. 

```{r}
set.seed(1)
train_index <- createDataPartition(data$Attrition , times =1, p = 0.7, list = FALSE)
train <- data[train_index,] 
test <- data[-train_index,]
```
5. **Modeling, Tuning & Evaluation**

##training control to tune 

```{r}
##random forest model 
control <- trainControl(method="repeatedcv", number=3, repeats=1)
random_forest_model <- train(dplyr::select(data, -Attrition), 
                             data$Attrition,
                             data=train, 
                             method="rf", 
                             preProcess="scale", 
                             trControl=control)

prediction_rfm <- predict(random_forest_model, newdata=select(test, -Attrition))
confusionMatrix(prediction_rfm,reference=test$Attrition,positive="Yes")

imp <- varImp(random_forest_model, scale=FALSE)
```
6. **Conclusion**
We can see that Salary has big impact in employees attrition
```{r}
plot(imp)
```


